---
title: "HW 1"
author: "Josh Winnes"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    theme: readable
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(ggthemes)
library(palmerpenguins)
```

# Section 1.2.5

## Exercise 1

```{r message=FALSE, warning=FALSE}
glimpse(penguins)
```

There are 344 rows and 8 columns in the penguins dataframe.

## Exercise 2

```{r message=FALSE, warning=FALSE}
?penguins
```

The `bill_depth_mm` column is "a number denoting bill depth (millimeters)".

## Exercise 3

```{r message=FALSE, warning=FALSE}
plot3 <- ggplot(penguins, aes(x = bill_length_mm, y=bill_depth_mm)) +
  geom_point() +
  labs(
    title = "Bill Depth (mm) vs. Bill Length (mm)",
    x = "Bill Length (mm)",
    y = "Bill Depth (mm)"
  )

plot3
```

This scatterplot shows that there is little correlation between bill length and bill depth though it looks like there are two clusters of points, possibly relating to species?

## Exercise 4

```{r message=FALSE, warning=FALSE}
plot4 <- ggplot(penguins, aes(x = bill_depth_mm, y=species)) +
  geom_point()

plot4
```

A better choice of geom would be `geom_boxplot()`.

## Exercise 5

```{raw}
ggplot(data = penguins) + 
  geom_point()
```

This gives an error because there is no aesthetic mapping. You would fix it by adding an aesthetic layer either globally in the `ggplot()` declaration or locally in the `geom_point()` layer.

## Exercise 6

According to the `?geom_point` page, "If `FALSE`, the default, missing values are removed with a warning. If `TRUE`, missing values are silently removed." For example:

```{r message=FALSE}
plot6 <- ggplot(penguins, aes(x = bill_length_mm, y=bill_depth_mm)) +
  geom_point(na.rm = TRUE) +
  labs(
    title = "Bill Depth (mm) vs. Bill Length (mm)",
    x = "Bill Length (mm)",
    y = "Bill Depth (mm)"
  )

plot6
```

## Exercise 7

```{r message=FALSE, warning=FALSE}
plot7 <- ggplot(penguins, aes(x = bill_length_mm, y=bill_depth_mm)) +
  geom_point(na.rm = TRUE) +
  labs(
    title = "Bill Depth (mm) vs. Bill Length (mm)",
    x = "Bill Length (mm)",
    y = "Bill Depth (mm)",
    caption = "Data come from the palmerpenguins package."
  )

plot7
```

## Exercise 8

```{r message=FALSE, warning=FALSE}
plot8 <- ggplot(penguins, aes(x=flipper_length_mm, y=body_mass_g))+
  geom_point(aes(color=bill_depth_mm))+
  geom_smooth()

plot8
```


# Section 1.4.3

## Exercise 1

```{r message=FALSE, warning=FALSE}
plot1 <- ggplot(penguins) +
  geom_bar(aes(y=species))

plot1
```

## Exercise 2

```{r message=FALSE, warning=TRUE}
ggplot(penguins, aes(x = species)) +
  geom_bar(color = "red")
```

```{r message=FALSE, warning=TRUE}
ggplot(penguins, aes(x = species)) +
  geom_bar(fill = "red")
```

They are the same chart except that the first one has a red border around the grey boxes while the second one has the boxes filled in with the color red. `fill` is far more useful for changing the color of the boxes.

# Section 1.5.5

## Exercise 1

```{r message=FALSE, warning=FALSE}
?mpg
mpg
```

The documentation honestly doesn't really give much help in finding the data types of the columns. However, running `mpg` is much more useful and you can see that `manufacturer`, `model `, `trans`, `drv`, `fl`, `year`, and `class` are categorical while `displ`, `cyl`, `cty`, and `hwy` are numerical. 

## Exercise 2

```{r message=FALSE, warning=FALSE}
ggplot(mpg, aes(x=displ, y=hwy))+
  geom_point(aes(color=model))


ggplot(mpg, aes(x=displ, y=hwy))+
  geom_point(aes(color=cty))

ggplot(mpg, aes(x=displ, y=hwy))+
  geom_point(aes(size=model))

ggplot(mpg, aes(x=displ, y=hwy))+
  geom_point(aes(size=cty))

ggplot(mpg, aes(x=displ, y=hwy))+
  geom_point(aes(shape=model))

# doesn't run because shape has to be assigned to a categorical variable
# ggplot(mpg, aes(x=displ, y=hwy))+
  # geom_point(aes(shape=cty))
```

The plots above show how `color`, `shape`, and `size` affect the scatterplot when they are mapped to categorical vs. continuous variables. In this case, there are so many models that the colors are hard to distinguish and there actually aren't enough shapes. Also, shapes cannot be mapped to a continuous variable because shapes are not continuous.

## Exercise 3

```{r message=FALSE, warning=FALSE}
ggplot(mpg)+
  geom_point(aes(x=hwy, y=displ, linewidth=cty))
```

It doesn't look like anything happens when you map a third variable to `linewidth` which makes sense because this is a scatterplot so there aren't actually any lines, just points.


# Section 11.2.1

## Exercise 1

```{r message=FALSE, warning=FALSE}
ggplot(mpg, aes(x=displ, y=hwy))+
  geom_point(aes(color=drv))+
  labs(
    title = "Highway MPG vs. Engine Displacement (L)",
    caption = "Data provided from the ggplot2 library.",
    x = "Engine Displacement (L)",
    y = "Highway MPG",
    drv = "Drive Train"
  )
```

## Exercise 2

```{r message=FALSE, warning=FALSE}
ggplot(mpg, aes(x=cty, y=hwy))+
  geom_point(aes(color=drv, shape=drv))+
  labs(
    x = "City MPG",
    y = "Highway MPG",
    color = "Type of drive train",
    shape = "Type of drive train"
  )
```

# Section 11.3.1

## Exercise 4

Just like other layers, the `aes()` argument in `geom_label()` can be filled with parameters like `color`, `fill`, `size`, etc. which control the appearance of the background box. 

# Section 11.4.6

## Exercise 3

### Part A

```{R}
presidential |>
  mutate(id = 33 + row_number()) |>
  ggplot(aes(x = start, y = id, color = party)) +
  geom_point() +
  geom_segment(aes(xend = end, yend = id)) +
  scale_color_manual(values = c(Republican = "#E81B23", Democratic = "#00AEF3"))
```

I'm not really sure what this question was asking.

### Part B

```{r message=FALSE, warning=FALSE}
presidential |>
  mutate(id = 33 + row_number()) |>
  ggplot(aes(x = start, y = id, color = party)) +
  geom_point() +
  geom_segment(aes(xend = end, yend = id)) +
  scale_color_manual(values = c(Republican = "#E81B23", Democratic = "#00AEF3"))
```

### Part C

```{r message=FALSE, warning=FALSE}
presidential |>
  mutate(id = 33 + row_number()) |>
  ggplot(aes(x = start, y = id, color = party)) +
  geom_point() +
  geom_segment(aes(xend = end, yend = id)) +
  scale_color_manual(values = c(Republican = "#E81B23", Democratic = "#00AEF3")) +
  geom_label(aes(label=name), size=3)
```

### Part D

```{r message=FALSE, warning=FALSE}
presidential |>
  mutate(id = 33 + row_number()) |>
  ggplot(aes(x = start, y = id, color = party)) +
  geom_point() +
  geom_segment(aes(xend = end, yend = id)) +
  scale_color_manual(values = c(Republican = "#E81B23", Democratic = "#00AEF3")) +
  geom_label(aes(label=name), size=3) +
  labs(
    x = "Year",
    y = "Presidential ID",
    title = "US Presidential Term Lengths",
    color = "Party"
  )
```

### Part E



