---
title: "Pre-Class Assignment 2"
author: "Josh Winnes"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
library(tidyverse)
library(palmerpenguins)
library(ggthemes)
```

# Part 1

### Question A

```{r message=FALSE, warning=FALSE}

ggplot(data=penguins, mapping = aes(x=flipper_length_mm, y=body_mass_g)) + 
  geom_point(
    mapping = aes(color=species)
  ) +
  geom_smooth(method = 'lm') +
  labs(
    x = "Flipper length (mm)",
    y = "Body mass (g)",
    title="Body mass and flipper length"
  ) +
  scale_color_colorblind()

```

### Question B

From the boxplot in 1.5.1, you can see that the Gentoo species seems to have a higher median body mass. The density plot agrees with this claim.

### Question C

One common coding mistake is to accidentally put the `+` at the beginning of a line instead of after a line, for example:

```{raw}
ggplot(penguins, aes(x=body_mass_g, y=flipper_length_mm))
+ geom_point()
```

Another common mistake is to forget to close parentheses and quotation marks. In order to get help with R functions, you can run `?function_name` in the console to get access to the help page.

# Part 2

I don't have any other questions from the reading at the moment.
