---
title: "Data Import Activity"
author: "Josh Winnes"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

# Getting Data

### Exercise 3

```{r data_import message=FALSE, warning=FALSE}
alone_2023 <- read_csv("https://www2.census.gov/programs-surveys/acs/data/2023/1_year_geographic_comparison_tables/GCT0802.csv", comment="TITLE:")
transit_2023 <- read_csv("https://www2.census.gov/programs-surveys/acs/data/2023/1_year_geographic_comparison_tables/GCT0804.csv", comment="TITLE:")

glimpse(alone_2023)
glimpse(transit_2023)
```
### Exercise 4 

I'm able to import the data fine, but it looks like the column names are a little messed up. I can tell what they're supposed to be, but it looks like there might be some subtitles that got put in as row 1. I added the `comment = 'TITLE:'` parameter in the `read_csv()` function to get rid of that.

### Exercise 5 

It looks like the data types are `<chr>` for `TBLID`, `GEOID`, `GEONAME`, and `MG_ERROR` while `PROFLN` and `ESTIMATE` are `<dbl>`. The only things that I think could be better would be to change `MG_ERROR` to `<dbl>` by assuming that they are all +/- and just putting the numeric value in.

# Renaming Columns

### Exercises 6 and 7

```{r renaming_columns message=FALSE, warning=FALSE}
# Exercise 6
library(janitor)

# Exercise 7
alone_2023 <- alone_2023 |>
  clean_names() |>
  mutate(commute_alone_pct = estimate) |>
  mutate(commute_alone_mg_error = -1*parse_number(mg_error))

transit_2023 <- transit_2023 |>
  clean_names() |>
  mutate(commute_transit_pct = estimate) |>
  mutate(commute_transit_mg_error = -1*parse_number(mg_error))
```

# Wrangling and Joining

### Exercise 8

```{r wrangling_and_joining message=FALSE, warning=FALSE}
alone_2023 <- alone_2023 %>% select(geoid, geoname, commute_alone_pct, commute_alone_mg_error)
transit_2023 <- transit_2023 %>% select(geoid, geoname, commute_transit_pct, commute_transit_mg_error)
```

# Past Data

### Exercise 9

```{r past_data message=FALSE, warning=FALSE}
url_alone_2019 <- "https://www2.census.gov/programs-surveys/acs/summary_file/2019/data/1_year_geographic_comparison_tables/GCT0802.csv"
url_transit_2019 <- "https://www2.census.gov/programs-surveys/acs/summary_file/2019/data/1_year_geographic_comparison_tables/GCT0804.csv"

alone_2019 <- read_csv(url_alone_2019, comment="TITLE:")
transit_2019 <- read_csv(url_transit_2019, comment="TITLE:")

alone_2019 |> 
  clean_names() |>
  mutate(commute_alone_pct = estimate) |>
  mutate(commute_alone_mg_error = -1*parse_number(mg_error))
alone_2019 <- alone_2019 %>% select(geoid, geoname, commute_alone_pct, commute_alone_mg_error)

transit_2019 |>
  clean_names() |>
  mutate(commute_transit_pct = estimate) |>
  mutate(commute_alone_mg_error = -1*parse_number(mg_error))
transit_2019 <- transit_2019 %>% select(geoid, geoname, commute_transit_pct, commute_transit_mg_error)
```

